**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-12 submission/manuscript.Rmd
now with references

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: plos.csl
---
\vspace{35mm}

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States (cite). CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen. These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (cite). The proximal and distal human colon are comprised of different microhabitats that likely enrich for different microbial membership at each site. Indeed, a number of studies have demonstrated that microbial populations differ throughout the gastrointestinal tract in accordance with oxygen, nutrient and pH gradients (cite). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish.

add something about sidedness of the gut - embryonic origins, different lymph and blood supplies - more background about male/female differences probably good too

Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (cite). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (cite). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (cite). These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_. Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (cite Niel).  However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site.

Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures (cite). While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel (cite). Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (Jalanka 2015, Harrell 2012). As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

>>>>

---
title: '**Machine learning identifies unique taxa differentiating proximal and distal
  human colonic microbiota**'
fontsize: 11pt
geometry: margin=1.0in
output:
  html_document: default
  pdf_document:
    includes: null
    keep_tex: yes
csl: plos.csl
bibliography: references.bib 
---
\vspace{35mm}

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. Tumors found in the proximal colon are typically flatter, sessile tumors that form along the inside of the colon wall whereas tumors of the distal colon are bulbous and project into the lumen. These features make cancers of the proximal colon less likely to present with bowel symptoms and thus are associated with an increase in morbidity and mortality (@Yamauchi2012). CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms. In contrast, right-sided colon tumors are bulky and project into the lumen, often causing no symptoms until they have developed into carcinomas (@Benedix2010). Because of the absence of symptoms, right-sided tumors, which comprise up to 47% of CRCs, have a significantly higher mortality rate (@Benedix2010). The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_. Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).  However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors. However, the contribution of the gut microbiota to these subtypes is largely undefined.

Furthermore, profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer (@Jalanka2014, @Harrell2012). As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
